---
title: "Other_Theft"
author: "Zaira Mae"
date: "2025-03-25"
output: html_document
---

```{r}
library(mgcv)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(spacetime)
library(sp)
library(animation)
library(gstat)
library(tidygeocoder)
library(maps)
library(lubridate)
library(reshape2)
library(sf)
library(corrplot)
library(pheatmap)
```

# Load the cleaned dataset

```{r}
Thft_vhcl <- read.csv("Theft_Vehicle_data.csv")
Thft_vhcl
```

Description of the data This data is about one crime type from the original dataset which is the Other Theft (Theft of property that includes personal items like purse, wallet, cellphone, laptop, etc., bicycle, etc.).

Neighborhood: The Vancouver Police Department uses Statistics Canada’s definition of neighbourhoods within municipalities. Neighbourhoods within the City of Vancouver are based on the census tract (CT) concept within the census metropolitan area (CMA).

Year: A four-digit field that indicates the year when the reported crime activity occurred.

Month: A numeric field that indicates the month when the reported crime activity occurred.

Total_Crimes: A total number of crimes that happened in that month from 2010 to 2024 per neighborhood.

# Loading the geojson file

```{r}
vancouver_geo <- st_read("local-area-boundary.geojson")
vancouver_geo
as.data.frame(st_coordinates(st_centroid(vancouver_geo$geometry)))
vancouver_geo$centroid <- st_centroid(vancouver_geo$geometry)
```

```{r}
colnames(Thft_vhcl)
```

```{r}
colnames(vancouver_geo)
```

```{r}
Thft_vhcl <- Thft_vhcl %>% 
   rename(name = Neighbourhood)
colnames(Thft_vhcl)
```

```{r}
Thft_vhcl <- Thft_vhcl %>%
  mutate(Time_Col = make_date(Year, 1))  

```

#Visualization from 2010-2024

```{r}
# Rename columns to match before joining
Thft_vhcl <- Thft_vhcl %>% rename()

# Perform the left join
Thft_vhcl_geo_combination <- left_join(vancouver_geo, Thft_vhcl, by = "name") %>%
  st_as_sf()

# Print the result
Thft_vhcl_geo_combination
```

```{r}
# Aggregate data by Neighborhood
visualization <- Thft_vhcl_geo_combination |> 
  group_by(name) |>  
  summarise(Total_Crimes = sum(Total_Crimes, na.rm = TRUE))

```

```{r}
# Compute label positions using st_point_on_surface() for better placement
Thft_vhcl_visualization <- visualization |> 
  mutate(label_position = st_point_on_surface(geometry))
# Create the map with neighborhood names and case counts
Thft_vhcl_visualization_plot <- ggplot(Thft_vhcl_visualization) +
  geom_sf(aes(fill = Total_Crimes), color = "black", lwd = 0.15) +
  scale_fill_viridis_c(option = "magma", direction = -1) +  
  theme_minimal() +

  # Add Neighborhood names inside the polygons
  geom_sf_text(aes(geometry = label_position, label = name), size = 3, color = "darkgreen") +  

  # Add Case Count inside polygons
  geom_sf_text(aes(geometry = label_position, label = Total_Crimes), size = 3, color = "maroon", nudge_y = 0.002)

# Display the map
Thft_vhcl_visualization_plot
ggsave("map.png", plot = Thft_vhcl_visualization_plot, width = 6, height = 4, dpi = 300)
```

```{r}
#Extract coordinates as a data frame
coords_df <- as.data.frame(st_coordinates(st_centroid(vancouver_geo$geometry)))

# Rename columns for clarity
colnames(coords_df) <- c("X", "Y")

```

```{r}
# First, get the neighborhood centroids with their names
neighborhood_centroids <- vancouver_geo %>%
  mutate(
    
    X = st_coordinates(st_centroid(geometry))[,1],
    Y = st_coordinates(st_centroid(geometry))[,2]
  ) %>%
  as.data.frame() %>%
  select(name, X, Y)

# Then merge with your crime data
combined_data <- Thft_vhcl %>%
  left_join(neighborhood_centroids, by = "name")
# Rename the X&Y to longitude and latitude
combined_data <- combined_data %>%
  rename(Latitude = Y)
combined_data <- combined_data %>%
  rename(Longitude = X)


# View the result
combined_data
  
```

```{r}
combined_data <- combined_data %>%
  rename(Neighborhood = name)
head(combined_data)
```

```{r}
combined_data$Neighborhood <- as.factor(combined_data$Neighborhood)
combined_data$Month<-as.factor(combined_data$Month)
```

```{r}
# Split the data into training (80%) and test (20%)
set.seed(123)  # For reproducibility

# Ensure you're using the correct dataframe name (combined_data, not crime)
train_indices <- sample(seq_len(nrow(combined_data)), size = 0.8 * nrow(combined_data))
train_data <- combined_data[train_indices, ]
test_data <- combined_data[-train_indices, ]

# Verify the split
cat("Training data rows:", nrow(train_data), "\n")
cat("Test data rows:", nrow(test_data), "\n")
```

```{r}
train_data$Month <- as.numeric(train_data$Month)
```

# --- Fit models ---

```{r}
Model1 <- gam(Total_Crimes ~ te(Latitude, Longitude, bs = c("tp", "tp"), k = c(12, 12)) +
                                s(Month, k = 12, bs = "cc"),
              data = train_data, method = "REML", family = nb(link = "log"))

```

```{r}
Model2 <- gam(Total_Crimes ~ te(Month, Longitude, bs = c("cc", "tp"), k = c(12, 12)),
              data = train_data, method = "REML", family = nb(link = "log"))
```

```{r}
Model3 <- gam(Total_Crimes ~ s(Longitude, k = 12, bs = "tp") +
                                s(Latitude, k = 12, bs = "tp") +
                                s(Month, k = 12, bs = "cc"),
              data = train_data, method = "REML", family = nb(link = "log"))
```

# --- Model list ---

```{r}
models <- list(Model1 = Model1, Model2 = Model2, Model3 = Model3)
```

# --- Gather evaluation metrics ---

```{r}
test_data$Month <- as.numeric(test_data$Month)
```

```{r}
library(Metrics)

model_results <- lapply(models, function(mdl) {
  s <- summary(mdl)
  pred <- predict(mdl, newdata = test_data, type = "response")
  data.frame(
    AIC = AIC(mdl),
    Deviance_Explained = round(s$dev.expl, 4),
    Adjusted_R2 = round(s$r.sq, 4),
    RMSE = round(rmse(test_data$Total_Crimes, pred), 4)
  )
})
```

# --- Combine into one table ---

```{r}
results_table <- bind_rows(model_results, .id = "Model")
print(results_table)
```

```{r}
anova.gam(Model1,Model2,Model3)
```

# Model 1 :

```{r}
Model1 <- gam(Total_Crimes ~ te(Latitude,Longitude, bs=c("tp","tp"), k=c(12,12)) + s(Month, k=12),data = train_data, method = "REML", family =nb(link = "log"))
summary(Model1)
```

#### Checking the Model

```{r}
gam.check(Model1)
```

#### Plot the model

```{r}
# Plot the model, including confidence intervals
plot(Model1)
```

```{r}
library(gratia)
draw(Model1)
```

```{r}
test_data$Month <- as.numeric(test_data$Month)
```

#### Prediction

```{r}
test_data$Predicted_Crimes_Model1 <- predict(Model1, newdata = test_data, type = "response")

```

```{r}
library(ggplot2)

# Define your palette (one color per neighborhood)
neighs <- unique(test_data$Neighborhood)
# you could also pick a palette function, e.g. scales::hue_pal()(length(neighs))
color_palette <- setNames(
  c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", 
    "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", 
    "#aec7e8", "#ffbb78", "#98df8a", "#ff9896", "#c5b0d5", 
    "#c49c94", "#f7b6d2", "#8c564b", "#e377c2", "#7f7f7f",
    "#bcbd22", "#17becf")[1:length(neighs)],
  neighs
)

for(nb in neighs) {
  df_nb <- subset(test_data, Neighborhood == nb)
  
  p <- ggplot(df_nb, aes(x = Month, y = Predicted_Crimes_Model1)) + 
    geom_line(color = color_palette[nb], size = 1) + 
    geom_ribbon(
      aes(
        ymin = Predicted_Crimes_Model1 * 0.9, 
        ymax = Predicted_Crimes_Model1 * 1.1
      ),
      fill = color_palette[nb],
      alpha = 0.2
    ) + 
    scale_x_continuous(breaks = c(1, 4, 7, 10)) + 
    labs(
      title = paste0("Crime Trend: ", nb),
      x     = "Month",
      y     = "Predicted Crimes (Model 1)"
    ) + 
    theme_minimal() + 
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title = element_text(face = "bold")
    )
  
  print(p)
}

```


# Model 3:
```{r}
Model3 <- gam(Total_Crimes ~ s(Longitude, k = 12, bs = "tp") +
                                s(Latitude, k = 12, bs = "tp") +
                                s(Month, k = 12, bs = "cc"),
              data = train_data, method = "REML", family = nb(link = "log"))
```

```{r}
summary(Model3)
```
### Checking the model

```{r}
gam.check(Model3)
```
### Plot the model

```{r}
plot(Model3)
```

```{r}
library(gratia)
draw(Model3)
```

### Prediction
```{r}
test_data$Month <- as.numeric(test_data$Month)
```


```{r}
test_data$Predicted_Crimes_Model3 <- predict(Model3, newdata = test_data, type = "response")
```

```{r}
library(ggplot2)

# Define your palette (one color per neighborhood)
neighs <- unique(test_data$Neighborhood)
# you could also pick a palette function, e.g. scales::hue_pal()(length(neighs))
color_palette <- setNames(
  c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", 
    "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", 
    "#aec7e8", "#ffbb78", "#98df8a", "#ff9896", "#c5b0d5", 
    "#c49c94", "#f7b6d2", "#8c564b", "#e377c2", "#7f7f7f",
    "#bcbd22", "#17becf")[1:length(neighs)],
  neighs
)

for(nb in neighs) {
  df_nb <- subset(test_data, Neighborhood == nb)
  
  p <- ggplot(df_nb, aes(x = Month, y = Predicted_Crimes_Model3)) + 
    geom_line(color = color_palette[nb], size = 1) + 
    geom_ribbon(
      aes(
        ymin = Predicted_Crimes_Model3 * 0.9, 
        ymax = Predicted_Crimes_Model3 * 1.1
      ),
      fill = color_palette[nb],
      alpha = 0.2
    ) + 
    scale_x_continuous(breaks = c(1, 4, 7, 10)) + 
    labs(
      title = paste0("Crime Trend: ", nb),
      x     = "Month",
      y     = "Predicted Crimes (Model 3)"
    ) + 
    theme_minimal() + 
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      axis.title = element_text(face = "bold")
    )
  
  print(p)
}


```

